Red Wine Quality by Hardeep Johar
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(reshape2)
library(tidyr)
library(dplyr)
library(GGally)
library(psych)
library(memisc)
library(gridExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv', header = T, row.names = 1, 
                 check.names = F)

```

**Summary**: We will explore 1,559 red wines with 11 variables on the
chemical properties of the wine. 3 wine experts rated the quality of each
wine, providing a rating between 0 (very bad) and 10 (excellent).

# Univariate Plots Section

Summary of the 11 variables we will be exploring.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
summary(wine)
```

Structure of the 11 variables we will be exploring.

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(wine)
```

Looks like we need to turn Quality into a factor.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Turn Quality into a factor
wine$quality <- as.factor(wine$quality)
str(wine$quality)
```

Great, now let's look at its frequency.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wine) +
  geom_histogram(stat ="count", color = "Green", fill = "yellow") +
  scale_x_discrete(name = "Quality Rating") +
  scale_y_continuous(name = "Count") +
  ggtitle("Histogram of Quality Rating") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0)) +
  theme(axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 12))

```

From above we can see Quality has a mean of 5.636 and a median of 6.
As expected, Quality works out to have a normal distribtion with more wines
having an average rating, with ratings decreasing in frequency towards the
extremities.

```{r echo=FALSE, message=FALSE, warning=FALSE}
tmpdata <- subset(wine, select = -c(quality))
w <- melt(tmpdata)
ggplot(w, aes(x = value)) + 
    facet_wrap(~variable, scales = "free") + 
    geom_histogram(color = "Blue", fill = "Orange") +
    labs(x = "", y = "Count") +
    ggtitle("Histograms of 11 Variables affecting Quality Rating") +
    theme(axis.title.x = element_text(face = "bold", colour = "#990000", 
                                      size = 8),
          axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
          axis.title.y = element_text(face = "bold", colour = "#990000",
                                      size = 8),
          axis.text.y = element_text(size = 8))

```

In order to better visualize data which has a long tail such as residual sugar, 
chlorides, free sulfer dioxide, total sulfer dioxide and sulphates, let's
do a log10 transformation.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Residual.sugar.1 <- ggplot(wine, aes(x = 1, y = residual.sugar)) + 
                      geom_jitter(alpha = 0.1) +
                      geom_boxplot(alpha = 0.2, color = 'red') +
                      labs(x = "", y = "Residual Sugar") +
                      ggtitle("Boxplot of Residual Sugar")
Residual.sugar.2 <- ggplot(wine, aes(x = residual.sugar)) + 
                      geom_histogram(bins = 30, color = "Blue",
                                     fill = "Orange") +
                      labs(x = "Residual Sugar (Log10)", y = "Count") +
                      ggtitle("Log 10 Histogram of Residual Sugar") +
                      scale_x_log10(breaks = c(.9, 2.2, 2.539, 15.5)) +
                      theme(axis.title.x = element_text(face = "bold", 
                                                        colour = "#990000", 
                                                        size = 8),
                      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                  size = 8),
                      axis.title.y = element_text(face = "bold", 
                                                  colour = "#990000", 
                                                  size = 8),
                      axis.text.y = element_text(size = 8))
grid.arrange(Residual.sugar.1, Residual.sugar.2, ncol = 2)
summary(wine$residual.sugar)
```

Residual Sugar has a mean of 2.200 and median of 2.539 g / dm^3. The above 
historgram has a normal distribtion. Outliers, as demonstrated in the box plot,
created a positively skewed distribiton and a long tail in the earlier 
historgram before the Log10 transformation was done.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Chlorides.1 <- ggplot(wine, aes(x = 1, y = chlorides)) + 
                      geom_jitter(alpha = 0.1) +
                      geom_boxplot(alpha = 0.2, color = 'red') +
                      labs(x = "", y = "Chlorides") +
                      ggtitle("Boxplot of Chlorides")
Chlorides.2 <- ggplot(wine, aes( x = chlorides)) + 
                      geom_histogram(bins = 30, color = "Blue", 
                                     fill = "Orange") +
                      labs(x = "Chlorides (Log10)", y = "Count") +
                      ggtitle("Log 10 Histogram of Chlorides") +
                      scale_x_log10(breaks = c(.012, .035, .079,
                                             .0875, .35, .611)) +
                      theme(axis.title.x = element_text(face = "bold", 
                                                        colour = "#990000", 
                                                        size = 8),
                      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                  size = 8),
                      axis.title.y = element_text(face = "bold", 
                                                  colour = "#990000", 
                                                  size = 8),
                      axis.text.y = element_text(size = 8))
grid.arrange(Chlorides.1, Chlorides.2, ncol = 2)
summary(wine$chlorides)
```

Chlorides have a mean of 0.087 and a median of0.079 g / dm^3. The above 
historgram has a normal distribtion. Outliers, as demonstrated in the box plot,
created a positively skewed distribiton and a long tail in the earlier 
historgram before the Log10 transformation was done.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Free.sulfur.dioxide.1 <- ggplot(wine, aes(x = 1, y = free.sulfur.dioxide)) + 
                      geom_jitter(alpha = 0.1) +
                      geom_boxplot(alpha = 0.2, color = 'red') +
                      labs(x = "", y = "Free Sulfer Dioxide") +
                      ggtitle("Boxplot of Free Sulfer Dioxide")
Free.sulfur.dioxide.2 <- ggplot(wine, aes(x = free.sulfur.dioxide)) + 
                      geom_histogram(bins = 30, color = "Blue",
                                     fill = "Orange") +
                      labs(x = "Free Sulfer Dioxide (Log10)", y = "Count") +
                      ggtitle("Log 10 Histogram of Free Sulfer Dioxide") +
                      scale_x_log10(breaks = c(1, 14, 15.87, 72)) +
                      theme(axis.title.x = element_text(face = "bold", 
                                                        colour = "#990000", 
                                                        size = 8),
                      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                  size = 8),
                      axis.title.y = element_text(face = "bold", 
                                                  colour = "#990000", 
                                                  size = 8),
                      axis.text.y = element_text(size = 8))
grid.arrange(Free.sulfur.dioxide.1, Free.sulfur.dioxide.2, ncol = 2)
summary(wine$free.sulfur.dioxide)
```

Free Sulfer Dioxide has a mean of 15.87 and a median of 14.00 mg /dm^3. The 
above historgram has a slightly bimodal distribtion. Outliers, as demonstrated
in the box plot, created a positively skewed distribiton and a long tail in the
earlier historgram before the Log10 transformation was done.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Total.sulfur.dioxide.1 <- ggplot(wine, aes(x = 1, y = total.sulfur.dioxide)) + 
                      geom_jitter(alpha = 0.1) +
                      geom_boxplot(alpha = 0.2, color = 'red' ) +
                      labs(x = "", y = "Total Sulfer Dioxide") +
                      ggtitle("Boxplot of Total Sulfer Dioxide")
Total.sulfur.dioxide.2 <- ggplot(wine, aes(x = total.sulfur.dioxide)) + 
                      geom_histogram(bins = 30, color = "Blue",
                                     fill = "Orange") +
                      labs(x = "Total Sulfer Dioxide (Log10)", y = "Count") +
                      ggtitle("Log 10 Histogram of Total Sulfer Dioxide") +
                      scale_x_log10(breaks = c(6, 38, 46.47, 28)) +
                      theme(axis.title.x = element_text(face = "bold", 
                                                        colour = "#990000", 
                                                        size = 8),
                      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                  size = 8),
                      axis.title.y = element_text(face = "bold", 
                                                  colour = "#990000", 
                                                  size = 8),
                      axis.text.y = element_text(size = 8))
grid.arrange(Total.sulfur.dioxide.1, Total.sulfur.dioxide.2, ncol = 2)
summary(wine$total.sulfur.dioxide)
```

Total Sulfer Dioxide has a mean of 46.47 and a median of 38.00 mg/dm^3. The 
above historgram has a normal distribtion. Outliers, as demonstrated in the 
box plot, created a positively skewed distribiton and a long tail in the 
earlier historgram before the Log10 transformation was done.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Sulphates.1 <- ggplot(wine, aes(x = 1, y = sulphates)) + 
                      geom_jitter(alpha = 0.1) +
                      geom_boxplot(alpha = 0.2, color = 'red') +
                      labs(x = "", y = "Sulphates") +
                      ggtitle("Boxplot of Sulphates")
Sulphates.2 <- ggplot(wine, aes(x = sulphates)) + 
                      geom_histogram(bins = 30, color = "Blue",
                                     fill = "Orange") +
                      labs(x = "Sulphates (Log10)", y = "Count") +
                      ggtitle("Log 10 Histogram of Sulphates") +
                      scale_x_log10(breaks = c(.33, .62, .6581, 2)) +
                      theme(axis.title.x = element_text(face = "bold", 
                                                        colour = "#990000", 
                                                        size = 8),
                      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                  size = 8),
                      axis.title.y = element_text(face = "bold", 
                                                  colour = "#990000", 
                                                  size = 8),
                      axis.text.y = element_text(size = 8))
grid.arrange(Sulphates.1, Sulphates.2, ncol = 2)
summary(wine$sulphates)
```

Sulphates have a mean of 0.6581 and a median of 0.6200 g/dm^3. The above
historgram has a normal distribtion. Outliers, as demonstrated in the box
plot, created a positively skewed distribiton and a long tail in the earlier
historgram before the Log10 transformation was done.

Let's take a closer look at Citric Acid.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(binwidth = .005, color = "Green", fill= "yellow") +
    labs(x = "Citric Acid", y = "Count") +
    ggtitle("Frequency Histogram of Citric Acid") +
    theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                    face = "bold", size = 22, hjust = 0)) +
    theme(axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                    face = "bold", size = 12))
summary(wine$citric.acid)
```

Looks like a slightly bimodal distribution with a mean of 0.271 and median of 
0.260. Why so many zeros? Does it have to do with wine quality? Let's look into 
this further in our bivariate analysis.

# Univariate Analysis

### What is the structure of your dataset?

There are 1,559 red wines with 11 variables affecting the quality of the wine 
(fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free
sulfur dioxide, total sulfur dioxide, density, pH, sulphates, and alcohol.) All
variable are numeric.  Quality was an integer and has been changed to a factor.

### What is/are the main feature(s) of interest in your dataset?

As we have yet to get into correlations, at this point the main interest is
the combination of many variables it takes to make a great wine. I would like to 
explore to what degree and with what weight each one influences wine quality. 
I would like to create a predictive model which can be used to rate wine.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The detailed correlations between variables will provide insight into which
variables have positive and negative effects on the quality score.

### Did you create any new variables from existing variables in the dataset?

So far, I did use the melt() function to make the above histograms more easily.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Some variables were normally distributed but others had long tails in which
case I did a log10 transformation in order to visualize the data better.

I did have to change "Quality" to a factor as well as remove it and use the 
melt() function to facet wrap all of the variables in one ggplot function.

# Bivariate Plots Section

\n Let's look at a Scatterplot Matrix to give us an high level overview of the
relationships.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
pairs.panels(wine)
```

Would like to look further into a few graphs which have outliers to make sure
nothing out of the ordinary is taking place.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = volatile.acidity, y = citric.acid), data = wine) +
  geom_point(alpha = .2) +
  labs(x = "Volatile Acidity", y = "Citric Acidity") +
  ggtitle("Volatile vs. Citric Acidity") +
  geom_smooth() +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0)) +
  theme(axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 12))
```

As expected, a few outliers with a high volatile acidity are causing the
graph to displace points further from the concentration of data points.  The
two variables have a correlation coefficient of -0.55.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = free.sulfur.dioxide), data = wine) +
  geom_point(alpha = .2) +
  labs(x = "Residual Sugar", y = "Free Sulfur Dioxide") +
  ggtitle("Residual Sugar vs. Free Sulfter Dioxide") +
  geom_smooth() +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0)) +
  theme(axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 12))
```

As expected, a few outliers with a high residual sugar are causing the
graph to displace points further from the concentration of data points. The
two variables have a correlation coefficient of 0.19.

The scatterplots confirmed there was no cause of concern as it was merely
data from outliers.

Let's take a closer look at the variables which have the top 3 highest 
correlation to Quality - Volatile Acidity, Citric Acid, and Sulphates so
we can get a better insight into what affects Quality the most.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, fill = quality), data = wine) +
  geom_histogram(binwidth = .01) +
  labs(x = "Volatile Acidity", y = "Count", color = "Quality Rating") +
  ggtitle("Volatile Acidity vs. Quality Rating") +
  scale_x_continuous(breaks = seq(.11,1.58,.7)) +
  geom_vline(aes(xintercept = mean(volatile.acidity)),
               color = "red", linetype = "dashed", size = .5) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666"))
```

As per my expectatations, the average amount of volatile acidity seems to be 
lower than the mean in higher quality wines because this would make them less
likely to taste like vinegar.  The two variables have a correlation coefficient
of -0.39.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, fill = quality), data = wine) +
  geom_histogram(binwidth = .01) +
  labs(x = "Citric Acid", y = "Count", color = "Quality Rating") +
  ggtitle("Citric Acid vs. Quality Rating") +
  scale_x_continuous(breaks = seq(0,1,.5)) +
  geom_vline(aes(xintercept = mean(citric.acid)),
               color = "red", linetype = "dashed", size = .5) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666"))
```

Seems many types of wines have 0 citric acid but definitely the higher rates
windes have a healthier dose of it.  This also met my expectations because
adding citric acid enhances the feeling of freshness to the taste buds. The
two variables have a correlation coefficient of 0.23.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates, fill = quality), data = wine) +
  geom_histogram(binwidth = .01) +
  labs(x = "Sulphates", y = "Count", color = "Quality Rating") +
  ggtitle("Sulphates vs. Quality Rating") +
  scale_x_continuous(breaks = seq(0.33,2,.5)) +
  geom_vline(aes(xintercept = mean(volatile.acidity)),
               color = "red", linetype = "dashed", size = .5) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666"))

```

Plotting sulphate count against quality yielded interesting results. On 
average, higher quality wines have a higher amount of sulphates. The
two variables have a correlation coefficient of 0.25.

This also met my expectations because in the past whenever I have an older
or better wine, it usually has a higher sulphate content.

The analysis lead me to wonder what do the best wines have more or less of so
I decided to look at the trends of the means of these three variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = volatile.acidity, color = quality), data = wine) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5) +
  scale_x_discrete(name = "Quality Rating") +
  labs(x = "Quality Rating", y = "Volatile Acidity") +
  ggtitle("Volatile Acidity Vs. Quality Rating") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.position = "none") +
  stat_summary(fun.y = "mean", 
               geom = "point",
               alpha = .4,
               color = "red", 
               shape = 8, 
               size = 4)
```

Better wines have lower volatile acidity.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = citric.acid, color = quality), data = wine) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5) +
  scale_x_discrete(name = "Quality Rating") +
  labs(x = "Quality Rating", y = "Citric Acid") +
  ggtitle("Citric Acid Vs. Quality Rating") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.position = "none") +
  stat_summary(fun.y = "mean", 
               geom = "point",
               alpha = .4,
               color = "red", 
               shape = 8, 
               size = 4)
```

Better wines have increasing citric acid.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = sulphates, color = quality), data = wine) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5) +
  scale_x_discrete(name = "Quality Rating") +
  labs(x = "Quality Rating", y = "Sulphates") +
  ggtitle("Sulphates Vs. Quality Rating") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.position = "none") +
  stat_summary(fun.y = "mean", 
               geom = "point",
               alpha = .4,
               color = "red", 
               shape = 8, 
               size = 4)
```

Better wines have more sulphates.

As you can see, all three of these yielded interesting results.  Better wines
clearly have: 

a.) lower volatile acidity (so they don't taste like Vinegar.)
b.) higher citric acid levels (giving them that refreshing taste); and 
c.) high levels of sulphates.

Additionally this lead me to think about the level of alcohol in good quality
wines.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = alcohol / 100, color = quality), data = wine) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5) +
  scale_x_discrete(name = "Quality Rating") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Quality Rating", y = "Alcohol") +
  ggtitle("Alcohol Vs. Quality Rating") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.position = "none") +
  stat_summary(fun.y = "mean", 
               geom = "point",
               alpha = .4,
               color = "red", 
               shape = 8, 
               size = 4)
```

Better wines have more alcohol! How interesting is that! If I hadn't seen this
I was always under the impression that they would be more balanced but this 
clearly shows the best wines have the higher alcohol levels.

# Bivariate Analysis

After looking at many variables correleted to each other and to quality,
it seems that there are numerous factors at play which lead to a good wine.
I look forward to the multivariate analysis.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

As discussed after the graphs, better wines have less volatile acidity, higher
citric acid and sulphate levels, and more alcohol.

After doing some additional research, the average amount of sulphates
increases as the quality of the wine increases in order to increase the
antimicrobial effects as better wines need to age for longer.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Relationships such as fixed acidity being highly correlated with pH, volatile
acidity being highly correlated with citric acid which is highly correlated
with pH and free sulfer dioxide being highly correlated with total sulfer
dioxide didn't surprise me as they are inherently related.

Another problem that I am noticing is that most of the wines are rated either
5 or 6 so the majority of the datapoints relate to them so I have to free the 
scales to adjust for this.

### What was the strongest relationship you found?
Fixed Acidity has a high correlation of .683 with pH and .672 with Citric Acid.
Volatile Acidity has a high correlation of .552 with Citric Acid.
Citric Acid has a high correlation of .542 with pH.
Free Sulfer Dioxide has a high correltion of .668 with Total Sulfer Dioxide.


# Multivariate Plots Section

Let's look at how the top 3 highest correlations to quality interact with one
another from a multivariate perspective.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x = sulphates, y = citric.acid, color = quality), data = wine) +
  geom_point(alpha = .8) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = "div") +
  labs(x = "Sulphates", y = "Citric Acid", color = "Quality Rating") +
  ggtitle("Sulphates & Citric Acid vs. Quality Rating") +
  coord_cartesian(xlim = c(quantile(wine$sulphates, 0), 
                           quantile(wine$sulphates, .99)), 
                  ylim = c(quantile(wine$citric.acid, 0), 
                           quantile(wine$citric.acid, .99))) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666")) +
  theme_dark()
```

Higher quality wines have a better balance of citric acid and sulphate levels.
Most interesting is that the lowest quality wines have the opposite (negative
slope) levels of sulphates and citric acid than the better quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = citric.acid, color = quality), 
       data = wine) +
  geom_point(alpha = .8) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = "div", palette = 2) +
  labs(x = "Volatile Acidity", y = "Citric Acid", color = "Quality Rating") +
  ggtitle("Volatile Acidity & Citric Acid vs. Quality Rating") +
  coord_cartesian(xlim = c(quantile(wine$volatile.acidity, 0), 
                           quantile(wine$volatile.acidity, .99)), 
                  ylim = c(quantile(wine$citric.acid, 0), 
                           quantile(wine$citric.acid, .99))) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666")) +
  theme_dark()

```

Here we can see that better wines have lower volatile acidity levels throughout
the spectrum of citric acid levels.  This makes sense because although citric
acid levels are higher in lower quality wines than volatile acidity is present
in higher quality wines. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = sulphates, color = quality), data = wine) +
  geom_point(alpha = .8) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = "div", palette = 4) +
  labs(x = "Volatile Acidity", y = "Sulphates", color = "Quality Rating") +
  ggtitle("Volatile Acidity & Sulphates vs. Quality Rating") +
  coord_cartesian(xlim = c(quantile(wine$volatile.acidity, 0), 
                           quantile(wine$volatile.acidity, .99)), 
                  ylim = c(quantile(wine$sulphates, 0), 
                           quantile(wine$sulphates, .99))) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666")) +
  theme_dark()
```

This further demonstrates the original results that there is lower volatile
acidity but higher sulphates in better wines.

Let's see if we can make a mathematical/linear model, see how much each variable
affects wine quality and see how accurate our model turns out to be.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#change wine quality back to numeric for linear modeling
wine$quality<-as.numeric(wine$quality)
#create linear model
x1 <- lm(wine$quality ~ wine$fixed.acidity)
x2 <- update(x1, ~. + wine$volatile.acidity)
x3 <- update(x2, ~. + wine$citric.acid)
x4 <- update(x3, ~. + wine$residual.sugar)
x5 <- update(x4, ~. + wine$chlorides)
x6 <- update(x5, ~. + wine$free.sulfur.dioxide)
x7 <- update(x6, ~. + wine$total.sulfur.dioxide)
x8 <- update(x7, ~. + wine$density)
x9 <- update(x8, ~. + wine$pH)
x10 <- update(x9, ~. + wine$sulphates)
x11  <- update(x10, ~. + wine$alcohol)
mtable(x1, x2, x3, x4, x5, x6, x7, x8, x9, x10, x11, sdigits = 5)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

As earlier thoughts and the three graphs above demonstrate:
Better wines have ore citric acid, alchohol and more sulphates with
less volatile acidity.

### Were there any interesting or surprising interactions between features?

I added facets of alochol buckets (10%, 11%, 12%, etc) but alochol content
seems to be medium to high for good wines - a bit less stark than alochol vs
quality showed when it was a bivariate analysis.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I did create a linear model.  Through this analysis I have learnt that no 1, 2,
or 3 variables make up a good wine.  The variables in the linear model only 
account for 36% of the variance.  This is after 11 variables contribute.  So the
limitation for the linear model is that other variables are accounting towards
the rating of the wine which are not included in the data set.  The strength
of the model is that we can see which factors do (although gradual) make a 
contribution towards accounting for the variance and we have accounted for 36%
of those factors.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# Reload the Data
wine <- read.csv('wineQualityReds.csv', header = T, row.names = 1, 
                 check.names = F)
# Turn Quality into a factor
wine$quality <- as.factor(wine$quality)

ggplot(aes(x = volatile.acidity, fill = quality), data = wine) +
  geom_histogram(binwidth = .01) +
  labs(x = "Volatile Acidity (g / dm^3)", y = "Count", 
       color = "Quality Rating") +
  ggtitle("Volatile Acidity vs. Quality Rating") +
  scale_x_continuous(breaks = seq(.11,1.58,.7)) +
  geom_vline(aes(xintercept = mean(volatile.acidity)),
               color = "red", linetype = "dashed", size = .5) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666"))
```

### Description One
This plot clearly shows that even though all wines have volatile acidity, there
is a clear trend towards better wines having less of it as observed by the
better rated wines being further from the mean.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = quality, y = alcohol / 100, color = quality), data = wine) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5) +
  scale_x_discrete(name = "Quality Rating") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Quality Rating", y = "Alcohol") +
  ggtitle("Alcohol Vs. Quality Rating") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.position = "none") +
  stat_summary(fun.y = "mean", 
               geom = "point",
               alpha = .4,
               color = "red", 
               shape = 8, 
               size = 4)
```

### Description Two
Higher quality wines in our sample do have higher percentages of alcohol.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = sulphates, y = citric.acid, color = quality), data = wine) +
  geom_point(alpha = .8) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = "div") +
  labs(x = "Sulphates (g / dm^3)", y = "Citric Acid (g / dm^3)",
       color = "Quality Rating") +
  ggtitle("Sulphates & Citric Acid vs. Quality Rating") +
  coord_cartesian(xlim = c(quantile(wine$sulphates, 0), 
                           quantile(wine$sulphates, .99)), 
                  ylim = c(quantile(wine$citric.acid, 0), 
                           quantile(wine$citric.acid, .99))) +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                  face = "bold", size = 22, hjust = 0),
       axis.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                 face = "bold", size = 12),
       legend.title = element_text(family = "Trebuchet MS", color = "#666666", 
                                   face = "bold", size = 12),
       legend.text = element_text(family = "Trebuchet MS", color = "#666666")) +
  theme_dark()
```

### Description Three
A nice way to visualize good wines have a healthy amount of sulphates as well
as citric acid.  Conversely you can also see that poor rated wines have the
opposite characteristics of each.

------

# Reflection

I started off hoping this dataset would be like the diamonds dataset, where
each "C" has a meaningful impact on the target (in that case, price) and in the
end accounted for almost 99% of the variance.

This dataset was very different.  It taught me how delicate wines really are.
I completely understand why there is a huge market for rare wines because
even with 11 variables above we could only account for ~36% of the variance!
This was very suprising to me and lead me to research the topic further.
Researching online shows that everything from climate to how you trim the grape
vines impacts wine quality.

Struggles: I definitely struggled to clean up the graphs but learned a lot in
the process.  Each layer in ggplot can be very powerful.  Additionally although
it might not seem like there was a lot of data wrangling, there was!  I cut
numerous variables only to revert back to the original dataset because I didn't
want to clutter this report.  Changing data types form numeric to factors 
for the analysis and then back to numeric for the linear model and then back to
factors for the remaining graphs keep me quite in touch with the structure of
the dataset and made me realize its importance.  I practiced using the melt() 
function as well as the gather() funciton and then stuck with melt() to combine
12 columns in 2 so that ggplot could easily create the histogram grids.
Changing axis fonts, labels, sizes taught me about using the theme funciton.
Adding colors and color scales was good practice and made the data stand out
better as well as made it more interesting visually.

What went well was that we did find numerous trends which impact a good quality
wine, such as citric acid and sulphate levels.  As mentioned above, a larger
dataset with additional characteristics would be fascinating to look at with
the goal of creating an improved linear model.